---
title: "Building a model with odin"
format:
  revealjs:
    preview-links: auto
execute:
  echo: true
  message: true
  output: true
  warning: true
---

```{r}
#| include: false
#| cache: false
set.seed(1)
source("common.R")
options(monty.progress = FALSE)
```

# `odin`

- A "domain specific language"
- Originally designed for ordinary differential equations
- Subsequently extended to difference equations (enabling discrete-time stochastic models)
- Refined into second version `odin2`

::: {.notes}
Odin is a DSL - a domain specific language. It exists to describe a specific problem efficiently. We developed it to describe ordinary differential equations which evolve a system in continuous time. This is a great target for DSL because it's just a collection of mathematical truths - rates exist independent of any idea of order of operation.
:::

# `odin`

```{r}
library(odin2)
library(dust2)
library(monty)
```

## ODE models {.smaller}

:::: {.columns}

::: {.column width="65%"}

::: {style="font-size: 1.5em"}

```{r}
#| echo: false
#| results: "asis"
r_output(readLines("models/sir-basic-ode.R"))
```

:::

:::

::: {.column width="35%"}
\begin{gather*}
\frac{dS}{dt} = -\beta S \frac{I}{N}\\
\frac{dI}{dt} = \beta S \frac{I}{N} - \gamma I\\
\frac{dR}{dt} = \gamma I
\end{gather*}
:::

::::

::: footer
In the book: [Getting started with odin](https://mrc-ide.github.io/odin-monty/odin.html)
:::

. . .

Things to note:

* out of order definition
* every variable has `initial` and `deriv` pair
* parameters declared with `parameter()`, default values can be given in brackets though not required


## Compiling the model with `odin2` {.smaller}

::: {style="font-size: 1.5em"}

```{r}
#| echo: false
#| results: "asis"
r_output(model_compile_code("sir_ode", "models/sir-basic-ode.R"))
```

```{r}
#| include: false
sir_ode <- odin("models/sir-basic-ode.R")
```

:::

<br> You can also compile code from a file by its path e.g. `odin("models/sir-basic.R")`.

::: {.notes}
The `odin` call generates C code, compiles it with `gcc` or `clang` to create a shared library, and loads that plus support code into R to create a generator object `gen`
:::

::: footer
In the book: [Getting started with odin](https://mrc-ide.github.io/odin-monty/odin.html)
:::

## Running the model with `dust2` {.smaller}

::: {style="font-size: 1.5em"}

```{r}
sys <- dust_system_create(sir_ode, pars = list())
dust_system_set_state_initial(sys)
t <- seq(0, 100)
y <- dust_system_simulate(sys, t)
```

:::

<br> The output has dimensions number of states x number of timepoints

::: {style="font-size: 1.5em"}

```{r}
dim(y)
```

:::

::: footer
In the book: [Getting started with odin](https://mrc-ide.github.io/odin-monty/odin.html)
:::

## Unpacking states {.smaller}

Output can be nicely unpacked into the different states using `dust_unpack_state`

::: {.panel-tabset}

### Code

::: {style="font-size: 1.5em"}

```r
y <- dust_unpack_state(sys, y)

plot(t, y$I, type = "l", xlab = "Time", ylab = "I")
```
:::

### Plot

```{r}
#| echo: false
y <- dust_unpack_state(sys, y)

plot(t, y$I, type = "l", xlab = "Time", ylab = "I")
```

:::

::: footer
In the book: [Getting started with odin](https://mrc-ide.github.io/odin-monty/odin.html)
:::


::: {.notes}
From the generator object `gen` we can construct a model (`mod`), here using no parameters -- just using default parameters as defined above. We run over a set of times and output the value of the system at a number of intermediate times.
:::


## Stochastic models {.smaller}

:::: {.columns}

::: {.column width="55%"}

::: {style="font-size: 1.5em"}

```{r}
#| echo: false
#| results: "asis"
r_output(readLines("models/sir-basic.R"))
```

:::

:::

::: {.column width="45%"}
\begin{gather*}
S(t + \Delta t) = S(t) - n_{SI}\\
I(t + \Delta t) = I(t) + n_{SI} - n_{IR}\\
R(t + \Delta t) = R(t) + n_{IR}
\end{gather*}

:::

::::

. . .

* `dt` is a special parameter
* every variable has `initial` and `update` pair

::: footer
In the book: [Stochasticity](https://mrc-ide.github.io/odin-monty/stochasticity.html)
:::


## ...compared with ODE models

:::: {.columns}

::: {.column style="font-size: 75%"}
```{r}
#| echo: false
#| results: "asis"
r_output(readLines("models/sir-basic.R"))
```
:::

::: {.column style="font-size: 75%"}
```{r}
#| echo: false
#| results: "asis"
r_output(readLines("models/sir-basic-ode.R"))
```

:::

::::

::: footer
In the book: [Stochasticity](https://mrc-ide.github.io/odin-monty/stochasticity.html)
:::

## Compiling with `odin2`

```{r}
#| echo: false
#| results: "asis"
r_output(model_compile_code("sir", "models/sir-basic.R"))
```

```{r}
#| include: false
sir <- odin("models/sir-basic.R")
```

::: footer
In the book: [Stochasticity](https://mrc-ide.github.io/odin-monty/stochasticity.html)
:::


## Functions in `odin2`

:::{}
* we have used the exponential function `exp()`
* many mathematical functions are supported in `odin2` - a full list can be found [here](https://mrc-ide.github.io/odin2/articles/functions.html)
:::

:::{}
* we have used `Binomial(S, p_SI)` for a Binomial random draw with `size = S` and `prob = p_SI`
* many distributions are supported - a full list with their names and parameterisations can be found [here](https://mrc-ide.github.io/odin2/articles/functions.html#distribution-functions)
:::
## `time` in `odin`

* `time` is a special (continuous) variable in `odin` models
* In discrete-time (stochastic) models, `time` is updated in increments of size `dt`


::: footer
In the book: [On the nature of time](https://mrc-ide.github.io/odin-monty/time.html)
:::

## Running a single simulation {.smaller}

::: {.panel-tabset}

#### Code

::: {style="font-size: 1.5em"}

```r
sys <- dust_system_create(sir, pars = list(), dt = 0.25)
dust_system_set_state_initial(sys)
t <- seq(0, 100)
y <- dust_system_simulate(sys, t)
y <- dust_unpack_state(sys, y)

plot(t, y$I, type = "l", xlab = "Time", ylab = "Infected population")
```

:::


#### Plot

```{r}
#| echo: false
sys <- dust_system_create(sir, pars = list(), dt = 0.25)
dust_system_set_state_initial(sys)
t <- seq(0, 100)
y <- dust_system_simulate(sys, t)
y <- dust_unpack_state(sys, y)

plot(t, y$I, type = "l", xlab = "Time", ylab = "Infected population")
```

:::

::: footer
In the book: [Stochasticity](https://mrc-ide.github.io/odin-monty/stochasticity.html)
:::

## Running multiple simulations {.smaller}

::: {.panel-tabset}

#### Code

::: {style="font-size: 1.5em"}

```r
sys <- dust_system_create(sir, pars = list(), n_particles = 50,
                                 dt = 0.25)
dust_system_set_state_initial(sys)
t <- seq(0, 100)
y <- dust_system_simulate(sys, t)
y <- dust_unpack_state(sys, y)
matplot(t, t(y$I), type = "l", lty = 1, col = "#00000044",
        xlab = "Time", ylab = "Infected population")
```

:::

#### Plot

```{r}
#| echo: false
sys <- dust_system_create(sir, pars = list(), n_particles = 50,
                                 dt = 0.25)
dust_system_set_state_initial(sys)
t <- seq(0, 100)
y <- dust_system_simulate(sys, t)
y <- dust_unpack_state(sys, y)
matplot(t, t(y$I), type = "l", lty = 1, col = "#00000044",
        xlab = "Time", ylab = "Infected population")
```

:::

::: footer
In the book: [Stochasticity](https://mrc-ide.github.io/odin-monty/stochasticity.html)
:::

## Calculating incidence with `zero_every`

```{r}
#| echo: false
#| results: "asis"
r_output(model_compile_code("sir", "models/sir-incidence.R"))
```

```{r}
#| include: false
sir <- odin("models/sir-incidence.R")
```

::: footer
In the book: [On the nature of time](https://mrc-ide.github.io/odin-monty/time.html)
:::

## Incidence accumulates then resets {.smaller}

::: {.panel-tabset}

#### Code

::: {style="font-size: 1.5em"}

```r
sys <- dust_system_create(sir, pars = list(), dt = 1 / 128)
dust_system_set_state_initial(sys)
t <- seq(0, 20, by = 1 / 128)
y <- dust_system_simulate(sys, t)
y <- dust_unpack_state(sys, y)

plot(t[t %% 1 == 0], y$incidence[t %% 1 == 0], type = "o", pch = 19,
     ylab = "Infection incidence", xlab = "Time")
lines(t, y$incidence, col = "red")
```

:::

#### Plot

```{r}
#| echo: false
sys <- dust_system_create(sir, pars = list(), dt = 1 / 128)
dust_system_set_state_initial(sys)
t <- seq(0, 20, by = 1 / 128)
y <- dust_system_simulate(sys, t)
y <- dust_unpack_state(sys, y)

plot(t[t %% 1 == 0], y$incidence[t %% 1 == 0], type = "o", pch = 19,
     ylab = "Infection incidence", xlab = "Time")
lines(t, y$incidence, col = "red")
```

:::

::: footer
In the book: [On the nature of time](https://mrc-ide.github.io/odin-monty/time.html)
:::

## Time-varying inputs: using `time`

```{r}
#| echo: false
#| results: "asis"
r_output(model_compile_code("sir", "models/sir-time.R"))
```

```{r}
#| include: false
sir <- odin("models/sir-time.R")
```

::: footer
In the book: [Inputs that vary over time](https://mrc-ide.github.io/odin-monty/interpolation.html)
:::


## Time-varying inputs: using `time` {.smaller}

::: {.panel-tabset}

#### Code

::: {style="font-size: 1.5em"}

```r
pars <- list(seed_time = 10, seed_size = 15)
sys <- dust_system_create(sir, pars = pars, dt = 0.25)
dust_system_set_state_initial(sys)
t <- seq(0, 100)
y <- dust_system_simulate(sys, t)
y <- dust_unpack_state(sys, y)

plot(t, y$I, type = "l", xlab = "Time", ylab = "Infected population")
```

:::

#### Plot

```{r}
#| echo: false
pars <- list(seed_time = 10, seed_size = 15)
sys <- dust_system_create(sir, pars = pars, dt = 0.25)
dust_system_set_state_initial(sys)
t <- seq(0, 100)
y <- dust_system_simulate(sys, t)
y <- dust_unpack_state(sys, y)

plot(t, y$I, type = "l", xlab = "Time", ylab = "Infected population")
```

:::

::: footer
In the book: [Inputs that vary over time](https://mrc-ide.github.io/odin-monty/interpolation.html)
:::


## Time-varying inputs: using `interpolate`

The `interpolate` function in odin can be used for time-varying parameters, with specification of

- the times of changepoints
- the values at those changepoints
- the type of interpolation: linear, constant or spline

::: footer
In the book: [Inputs that vary over time](https://mrc-ide.github.io/odin-monty/interpolation.html)
:::

## Time-varying inputs: using `interpolate`

```{r}
#| echo: false
#| results: "asis"
r_output(model_compile_code("sir", "models/sir-interpolate-constant.R"))
```

```{r}
#| include: false
sir <- odin("models/sir-interpolate-constant.R")
```

::: footer
In the book: [Inputs that vary over time](https://mrc-ide.github.io/odin-monty/interpolation.html)
:::

## Time-varying inputs: using `interpolate` {.smaller}

::: {.panel-tabset}

#### Code

::: {style="font-size: 1.5em"}

```r
pars <- list(beta_time = c(0, 30), beta_value = c(0.2, 1))
sys <- dust_system_create(sir, pars = pars, dt = 0.25)
dust_system_set_state_initial(sys)
t <- seq(0, 100)
y <- dust_system_simulate(sys, t)
y <- dust_unpack_state(sys, y)

plot(t, y$I, type = "l", xlab = "Time", ylab = "Infected population")
abline(v = 30, lty = 3)
```

:::

#### Plot

```{r}
#| echo: false
pars <- list(beta_time = c(0, 30), beta_value = c(0.2, 1))
sys <- dust_system_create(sir, pars = pars, dt = 0.25)
dust_system_set_state_initial(sys)
t <- seq(0, 100)
y <- dust_system_simulate(sys, t)
y <- dust_unpack_state(sys, y)

plot(t, y$I, type = "l", xlab = "Time", ylab = "Infected population")
abline(v = 30, lty = 3)
```

:::

::: footer
In the book: [Inputs that vary over time](https://mrc-ide.github.io/odin-monty/interpolation.html)
:::

## Using arrays

```{r}
#| echo: false
#| results: "asis"
r_output(model_compile_code("sir_age", "models/sir-age.R"))
```

```{r}
#| include: false
sir_age <- odin("models/sir-age.R")
```

::: footer
In the book: [Arrays](https://mrc-ide.github.io/odin-monty/arrays.html)
:::

## Using arrays {.smaller}

::: {style="font-size: 1.5em"}

```{r}
pars <- list(S0 = c(990, 1000),
             I0 = c(10, 0),
             m = matrix(c(1.8, 0.4, 0.4, 1.2) / 2000, 2, 2),
             beta = 0.2,
             gamma = 0.1,
             n_age = 2)
```

:::

<br> Here `m[i, j]` is the mean number of contacts per time unit for an individual in group `i` has with an individual in group `j`

::: footer
In the book: [Arrays](https://mrc-ide.github.io/odin-monty/arrays.html)
:::

## Using arrays {.smaller}

::: {.panel-tabset}

#### Code

::: {style="font-size: 1.5em"}

```r
sys <- dust_system_create(sir_age, pars = pars, dt = 0.25)
dust_system_set_state_initial(sys)
t <- seq(0, 100)
y <- dust_system_simulate(sys, t)
y <- dust_unpack_state(sys, y)

matplot(t, t(y$I), type = "l", lty = 1, col = c("red", "blue"),
        xlab = "Time", ylab = "Infected population")
legend("topright", c("children", "adults"), col = c("red", "blue"), lty = 1)
```

:::

#### Plot

```{r}
#| echo: false
sys <- dust_system_create(sir_age, pars = pars, dt = 0.25)
dust_system_set_state_initial(sys)
t <- seq(0, 100)
y <- dust_system_simulate(sys, t)
y <- dust_unpack_state(sys, y)

matplot(t, t(y$I), type = "l", lty = 1, col = c("red", "blue"),
        xlab = "Time", ylab = "Infected population")
legend("topright", c("children", "adults"), col = c("red", "blue"), lty = 1)
```

:::

::: footer
In the book: [Arrays](https://mrc-ide.github.io/odin-monty/arrays.html)
:::


## Key features of arrays

- All arrays (whether state variable or parameter) need a `dim` equation
- No use of `i`, `j` etc on LHS - indexing on the LHS is implicit
- Support for up to 8 dimensions, with index variables `i`, `j`, `k`, `l`, `i5`, `i6`, `i7`, `i8`
- Functions for reducing arrays such as `sum`, `prod`, `min`, `max` - can be applied over entire array or slices

::: footer
In the book: [Arrays](https://mrc-ide.github.io/odin-monty/arrays.html)
:::

## Arrays: model with age and vaccine

```{r}
#| echo: false
#| results: "asis"
r_output(model_compile_code("sir_age_vax", "models/sir-age-vax.R"))
```

::: footer
In the book: [Arrays](https://mrc-ide.github.io/odin-monty/arrays.html)
:::

## Arrays: boundary conditions {.smaller}

Multiline equations can be used to deal with boundary conditions, e.g. we have

::: {style="font-size: 1.5em"}

```r
new_S[, 1] <- S[i, j] - n_SI[i, j] - n_S_vax[i, j] + n_S_vax[i, n_vax]
new_S[, 2:n_vax] <- S[i, j] - n_SI[i, j] - n_S_vax[i, j] + n_S_vax[i, j - 1]
```

:::

which we could also write as

::: {style="font-size: 1.5em"}

```r
new_S[, ] <- S[i, j] - n_SI[i, j] - n_S_vax[i, j]
new_S[, 1] <- new_S[i, j] + n_S_vax[i, n_vax]
new_S[, 2:n_vax] <- new_S[i, j] + n_S_vax[i, j - 1]
```

:::

or another way of writing this would be to use `if ... else`

::: {style="font-size: 1.5em"}

```r
new_S[, ] <- S[i, j] - n_SI[i, j] - n_S_vax[i, j] +
    (if (j == 1) n_S_vax[i, n_vax] else n_S_vax[i, j - 1])
```

:::

::: footer
In the book: [Arrays](https://mrc-ide.github.io/odin-monty/arrays.html)
:::


## Dimensions {.smaller}

Dimensions can be hard-coded:

::: {style="font-size: 1.5em"}

```r
dim(S) <- 2
```

:::

or can be soft-coded via parameters:

::: {style="font-size: 1.5em"}

```r
dim(S) <- n_age
n_age <- parameter()
```

:::

::: footer
In the book: [Arrays](https://mrc-ide.github.io/odin-monty/arrays.html)
:::

## Dimensions {.smaller}

Array parameters can have dimensions explicitly declared

::: {style="font-size: 1.5em"}

```r
m <- parameter()
dim(m) <- c(n_age, n_age)
n_age <- parameter()
```

:::

or detected when the parameters are passed into the system - but a `dim` equation is needed where you must explicitly state the rank:

::: {style="font-size: 1.5em"}

```r
dim(m) <- parameter(rank = 2)
```

:::

::: footer
In the book: [Arrays](https://mrc-ide.github.io/odin-monty/arrays.html)
:::

## Dimensions {.smaller}

Dimensions of one array can be declared as the same as the dimensions of another:

::: {style="font-size: 1.5em"}

```r
dim(m) <- c(n_age, n_age)
dim(s_ij) <- dim(m)
```

:::

or dimensions of arrays that have the same dimensions can be declared collectively:

::: {style="font-size: 1.5em"}

```r
dim(m, s_ij) <- c(n_age, n_age)
```

:::

::: footer
In the book: [Arrays](https://mrc-ide.github.io/odin-monty/arrays.html)
:::

## Compilation error messages {.smaller}

Let's revisit the stochastic SIR model, which compiles successfully:

::: {style="font-size: 1.5em"}

```{r}
sir <- odin({
  update(S) <- S - n_SI
  update(I) <- I + n_SI - n_IR
  update(R) <- R + n_IR
  update(incidence) <- incidence + n_SI

  p_SI <- 1 - exp(-beta * I / N * dt)
  p_IR <- 1 - exp(-gamma * dt)
  n_SI <- Binomial(S, p_SI)
  n_IR <- Binomial(I, p_IR)

  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(incidence, zero_every = 1) <- 0

  N <- parameter(1000)
  I0 <- parameter(10)
  beta <- parameter(0.2)
  gamma <- parameter(0.1)
})

sir
```

:::

::: footer
In the book: [Debugging](https://mrc-ide.github.io/odin-monty/debugging.html)
:::

## Compilation error messages {.smaller}
Now let's remove the `initial(R)` equation

::: {style="font-size: 1.5em"}

```{r}
#| error: true

sir <- odin({
  update(S) <- S - n_SI
  update(I) <- I + n_SI - n_IR
  update(R) <- R + n_IR
  update(incidence) <- incidence + n_SI

  p_SI <- 1 - exp(-beta * I / N * dt)
  p_IR <- 1 - exp(-gamma * dt)
  n_SI <- Binomial(S, p_SI)
  n_IR <- Binomial(I, p_IR)

  initial(S) <- N - I0
  initial(I) <- I0
  initial(incidence, zero_every = 1) <- 0

  N <- parameter(1000)
  I0 <- parameter(10)
  beta <- parameter(0.2)
  gamma <- parameter(0.1)
})
```

:::

::: footer
In the book: [Debugging](https://mrc-ide.github.io/odin-monty/debugging.html)
:::

## Compilation error messages {.smaller}

We can run `odin_error_explain()` with the corresponding error code as suggested (clickable in RStudio) for more information

::: {style="font-size: 1.5em"}

```{r}
odin_error_explain("E2003")
```

:::

::: footer
In the book: [Debugging](https://mrc-ide.github.io/odin-monty/debugging.html)
:::

## Debugging {.smaller}

Here's a model that compiles but has a bug

::: {style="font-size: 1.5em"}

```{r}
sir <- odin({
  update(S) <- S - n_SI
  update(I) <- I + n_SI
  update(R) <- R + n_IR
  update(incidence) <- incidence + n_SI

  p_SI <- 1 - exp(-beta * I / N * dt)
  p_IR <- 1 - exp(-gamma * dt)
  n_SI <- Binomial(S, p_SI)
  n_IR <- Binomial(I, p_IR)

  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(incidence, zero_every = 1) <- 0

  N <- parameter(1000)
  I0 <- parameter(10)
  beta <- parameter(0.2)
  gamma <- parameter(0.1)
})

sir
```

:::

::: footer
In the book: [Debugging](https://mrc-ide.github.io/odin-monty/debugging.html)
:::

## Debugging {.smaller}

Let's run it

::: {style="font-size: 1.5em"}

```{r}
sys <- dust_system_create(sir)
dust_system_set_state_initial(sys)
t <- 50
res <- dust_system_simulate(sys, t)
dust_unpack_state(sys, res)
```

:::

The number of individuals in the `R` compartment has exceeded `N`!

::: footer
In the book: [Debugging](https://mrc-ide.github.io/odin-monty/debugging.html)
:::


## Debugging

You can stare at the code and hope to spot what's wrong, but that can be like trying to find a needle in a haystack, especially for larger models. 

So there are tools to help you:

* You can `print()` the values of variables in your model as it runs
* You can use the `browser()` function in the same way as you would in R.

::: footer
In the book: [Debugging](https://mrc-ide.github.io/odin-monty/debugging.html)
:::


## Debugging: using `print()` {.smaller}

With `print()` we can tell it what to print and (optionally) conditions for when to print.

::: {style="font-size: 1.5em"}

```{r}
sir <- odin({
  update(S) <- S - n_SI
  update(I) <- I + n_SI
  update(R) <- R + n_IR
  update(incidence) <- incidence + n_SI

  p_SI <- 1 - exp(-beta * I / N * dt)
  p_IR <- 1 - exp(-gamma * dt)
  n_SI <- Binomial(S, p_SI)
  n_IR <- Binomial(I, p_IR)

  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(incidence, zero_every = 1) <- 0

  N <- parameter(1000)
  I0 <- parameter(10)
  beta <- parameter(0.2)
  gamma <- parameter(0.1)
  
  print("S + I: {S + I}, R: {R}", when = S + I + R > N && time > 5)
})
```

:::

::: footer
In the book: [Debugging](https://mrc-ide.github.io/odin-monty/debugging.html)
:::


## Debugging: using `print()` {.smaller}

Then when we run it prints what we wanted with the times in square brackets:

::: {style="font-size: 1.5em"}

```{r}
sys <- dust_system_create(sir)
dust_system_set_state_initial(sys)
dust_system_run_to_time(sys, 10)
```

:::

<br> This may help us spot that we're not removing individuals from `I` when they move to `R`!

::: footer
In the book: [Debugging](https://mrc-ide.github.io/odin-monty/debugging.html)
:::


## Debugging: using `browser()` {.smaller}

With the `browser()` we need to declare the phase for inserting (`deriv` for ODE models, `update` for stochastic models) and optionally `when` to trigger it.

::: {style="font-size: 1.5em"}

```{r}
sir <- odin({
  update(S) <- S - n_SI
  update(I) <- I + n_SI
  update(R) <- R + n_IR
  update(incidence) <- incidence + n_SI

  p_SI <- 1 - exp(-beta * I / N * dt)
  p_IR <- 1 - exp(-gamma * dt)
  n_SI <- Binomial(S, p_SI)
  n_IR <- Binomial(I, p_IR)

  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(incidence, zero_every = 1) <- 0

  N <- parameter(1000)
  I0 <- parameter(10)
  beta <- parameter(0.2)
  gamma <- parameter(0.1)
  
  browser(phase = "update", when = S + I + R > N)
})
```

:::

::: footer
In the book: [Debugging](https://mrc-ide.github.io/odin-monty/debugging.html)
:::


## Debugging: using `browser()` {.smaller}

When we try to run it will pause during evaluation:

::: {style="font-size: 1.5em"}

```r
sys <- dust_system_create(sir)
dust_system_set_state_initial(sys)
dust_system_run_to_time(sys, 50)
ℹ dust debug ('update'; time = 2): see dust_browser for help
Called from: eval(substitute(expr), data, enclos = parent.frame())
Browse[1]> 
```

:::

<br> You can list all the variables

::: {style="font-size: 1.5em"}

```r
Browse[1]> ls()
 [1] "beta"      "gamma"     "I"         "I0"        "incidence" "N"         "n_IR"      "n_SI"      "p_IR"     
[10] "p_SI"      "R"         "S"         "time"  
```

:::

::: footer
In the book: [Debugging](https://mrc-ide.github.io/odin-monty/debugging.html)
:::

## Debugging: using `browser()` {.smaller}

You can inspect values and perform calculations

::: {style="font-size: 1.5em"}

```r
Browse[1]> N
[1] 1000
Browse[1]> I
[1] 18
Browse[1]> S
[1] 982
Browse[1]> I / N
[1] 0.018
```

:::

::: footer
In the book: [Debugging](https://mrc-ide.github.io/odin-monty/debugging.html)
:::

## Debugging: using `browser()`

- Enter `c` or `n`, to continue until the next time the `browser()` is triggered  
- If your model has a variable called `n`, you would need to type `message(n)` to see its value
- Enter `Q` to immediately exit to the console with an error.
- Run `dust_browser_continue()` to cancel all future `browser()` calls and continue running the model

::: footer
In the book: [Debugging](https://mrc-ide.github.io/odin-monty/debugging.html)
:::


## Next steps

- Get started [fitting `odin` models with `monty`](https://mrc-ide.github.io/odin-monty/slides/fitting.html)
- [Learn how to use `delay()`](https://mrc-ide.github.io/odin-monty/delays.html) for delay differential equations
- Understand the [order of events for discrete-time stochastic models](https://mrc-ide.github.io/odin-monty/order.html)
- For larger odin models, [learn about packaging them]((https://mrc-ide.github.io/odin-monty/packaging.html))
