# Debugging {#sec-debugging}

```{r}
#| include: false
source("common.R")
```

Odin provides some special functionality to help you debug your model:-
* You can `print()` the values of variables in your model as it runs
* You can use the `browser()` function in the same way as you would in R.
* You can examine the code generated from your model with `odin_show()`.


```{r}
library(odin2)
library(dust2)
```

## Using `print()`

In the code below, we can print out the value of `x` as we go.


```{r}
gen <- odin({
  update(x) <- Normal(x, 1)
  initial(x) <- 1
  print("x: {x}")
})
sys <- dust_system_create(gen())
dust_system_run_to_time(sys, 10)
```

Here we see the value of x for all of our 10 time points. If we want to be
more selective, we can choose when we would like to print using the `when`
argument to `print()` like this:


```{r}
gen <- odin({
  update(x) <- Normal(x, 1)
  initial(x) <- 1
  print("x above 1: {x}", when = x > 1)
})
sys <- dust_system_create(gen())
dust_system_run_to_time(sys, 10)
```

The `when` function can take more complicated logical expressions, including
brackets, and `||` and `&&` for `or` and `and` respectively. While at present
we can only have one `print()` line within your model, you can print multiple
variables togther like this:

```{r}
gen <- odin({
  update(x) <- Normal(x, 1)
  initial(x) <- 1
  update(y) <- Normal(x, 2)
  initial(y) <- 1
  print("x above 1: {x}, y = {y}", when = x > 1)
})
sys <- dust_system_create(gen())
dust_system_run_to_time(sys, 10)
```

### Formatting

We use [glue](https://glue.tidyverse.org/) to drive the formatting, which may
feel familiar to you already. Above, we surrounded variables with 
`{curly braces}` to print their value instead of their name, where all other
text was printed verbatim. 

You can control the precision of numbers being printed in the form below:

```{r}
gen <- odin({
  update(x) <- Normal(x, 1)
  initial(x) <- 1
  print("x: {x; .2f}")
})
sys <- dust_system_create(gen())
dust_system_run_to_time(sys, 10)
```

Here, the `.2f` gives us two decimal places; you may recognise that format 
if you've used `sprintf()`. By default, variables are printed as floats (`f`),
and you may sometimes find `g` useful for particularly small or large values.

Note that this is an experimental; see the debugging vignettes for current
limitations. 


## Using the interactive debugger, `browser()`

You can use the `browser()` function to ask odin to drop you into a debugger
which will appear very similar to doing so in R, but has a couple of arguments
that are specific to odin. 

* `phase`: the system phase where the debugger should be inserted; this will typically be `update` or `deriv`.
* `when`: optionally a condition that should be satisfied for the debugger to be triggered. You will typically want to set this or it will be called at every step.

For example, suppose we are interested in what happens at the end of an SIR
model run, when in the SIR model from `vignette("odin")` we might write:

```{r}
gen <- odin({
  p_IR <- 1 - exp(-gamma * dt)
  N <- parameter(1000)
  p_SI <- 1 - exp(-(beta * I / N * dt))
  n_SI <- Binomial(S, p_SI)
  n_IR <- Binomial(I, p_IR)
  update(S) <- S - n_SI
  update(I) <- I + n_SI - n_IR
  update(R) <- R + n_IR
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  beta <- parameter(0.2)
  gamma <- parameter(0.1)
  I0 <- parameter(10)

  browser(phase = "update", when = I < 10 && time > 20)
})
```

The location of the call to `browser()` does not matter; it will be activated at
the **end** of the phase.  The condition here might be something we cook up to
look at what happens as the number of individuals in the infected class starts
tailing off at the end of the simulation.

We create and initialise the system as normal:

```{r}
sys <- dust_system_create(gen())
dust_system_set_state_initial(sys)
```

However, when you run the system you will pause part way through evaluation:

```r
dust_system_run_to_time(sys, 200)
ℹ dust browser ('update'; time = 117): see `?dust_browser()` for help
Called from: eval(substitute(expr), data, enclos = parent.frame())
Browse[1]>
```

Now you can explore things that odin and dust know about, in a similar way
that you would in R.

```r
Browse[1]> ls()
 [1] "beta"  "gamma" "I"     "I0"    "N"     "n_IR"  "n_SI"  "p_IR"  "p_SI"
[10] "R"     "S"     "time"
```

and you can inspect values or perform calculations:

```r
Browse[1]> N
[1] 1000
Browse[1]> I
[1] 9
Browse[1]> S
[1] 178
Browse[1]> I / S
[1] 0.0505618
```

You can change values for testing just within this browser environment, but
note that the changes will *not* be injected back into your model.

If you press `c` or `n`, then odin the system will proceed to the next step and
drop you back into the debugger, when it reaches the browser command again,
just like in R. So if your model had a variable called `n`, you would have to
type `message(n)`, or use `sprintf` to see its value. You can exit with `Q
which will return you to the console with an error.  

To cancel all future browser calls and run to the end of your simulation:- 

```r
dust_browser_continue()
```

and then the next time you press `c` to continue, your simulation will run
to completion. 

## Showing the generated code

Sometimes just looking at the generated code can be helpful.  You can do this
with `odin_show`:

```{r}
odin_show({
  initial(x) <- 2
  update(x) <- Normal(x, 3)
})
```

Above may seem quite a bit of code, but if you look for the `initial` function
and the `update` function at the bottom you will see the code representing your
two model lines. To simplify things a little, we can show just a particular
method (usually `update` or `rhs`), with the `what` argument, for example:

```{r}
odin_show({
  initial(x) <- 2
  update(x) <- Normal(x, 3)
}, what = "update")
```
