# The monty DSL {#sec-monty-dsl}

```{r}
#| include: false
source("common.R")
```

The `monty` DSL provides a more intuitive way to define statistical models with `monty`. It is currently relatively basic and focuses on providing support for defining priors in Bayesian models. It fully support differentiability allowing to use gradient based samplers on these models.

```{r}
library(monty)
```

## A simple example

In chapter 4 of [Statistical Rethinking](https://xcelab.net/rm/), we build a regression model of height with parameters $\alpha$, $\beta$ and $\sigma$.  We can define the model for the prior probability of this model in monty by running

```{r}
prior <- monty_dsl({
  alpha ~ Normal(178, 20)
  beta ~ Normal(0, 10)
  sigma ~ Uniform(0, 50)
})
```

This will define a new `monty_model()` object that represents the prior, but with all the bits that we might need depending on how we want to use it:

We have model parameters

```{r}
prior$parameters
```

We can compute the domain for your model:

```{r}
prior$domain
```

We can draw samples from the model if we provide a `monty_rng` object

```{r}
rng <- monty_rng_create()
theta <- monty_model_direct_sample(prior, rng)
theta
```

We can compute the (log) density at a point in parameter space

```{r}
prior$density(theta)
```

The computed properties for the model are:

```{r}
prior$properties
```

## Distribution functions

In the above example we use distribution functions for the normal and uniform distributions. The distribution functions available for the `monty` DSL are the same as those for the `odin` DSL, the full list of which can be found [here](https://mrc-ide.github.io/odin2/articles/functions.html#distribution-functions).

## Dependent distributions

It is possible within the DSL to have the distribution of parameters to depend upon the value of other parameters:

```{r}
m <- monty_dsl({
    a ~ Normal(0, 1)
    b ~ Normal(a, 1)
})
```

This is particularly useful for the implementation of [hyperpriors](https://en.wikipedia.org/wiki/Hyperprior) when using the DSL to define priors.

Order of equations is important when using dependent distributions in the `monty` DSL! You cannot have the distribution of a parameter depend upon a parameter that is defined later. Thus rewriting the above example as

```
m <- monty_dsl({
    b ~ Normal(a, 1)
    a ~ Normal(0, 1)
})
```
would produce an error.

## Calculations in the DSL

Sometimes it will be useful to perform calculations in the code; you can do this with assignments.  Most trivially, giving names to numbers may help make code more understandable:

```{r}
m <- monty_dsl({
  mu <- 10
  sd <- 2
  a ~ Normal(mu, sd)
})
```

You can also use this to do things like:

```{r}
m <- monty_dsl({
  a ~ Normal(0, 1)
  b ~ Normal(0, 1)
  mu <- (a + b) / 2
  c ~ Normal(mu, 1)
})
```

Where `c` is drawn from a normal distribution with a mean that is the average of `a` and `b`.

## Pass in fixed data

You can also pass in a list of data with values that should be available in the DSL code.  For example, our first example:

```{r}
prior <- monty_dsl({
  alpha ~ Normal(178, 20)
  beta ~ Normal(0, 10)
  sigma ~ Uniform(0, 50)
})
```

Might be written as

```{r}
fixed <- list(alpha_mean = 170, alpha_sd = 20,
              beta_mean = 0, beta_sd = 10,
              sigma_max = 50)
prior <- monty_dsl({
  alpha ~ Normal(alpha_mean, alpha_sd)
  beta ~ Normal(beta_mean, beta_sd)
  sigma ~ Uniform(0, sigma_max)
}, fixed = fixed)
```

Values you pass in this way are **fixed** (hence the name!) and cannot be modified after the model object is created.

## Arrays

The `monty` DSL supports arrays with a similar syntax to `odin`. In this example

```{r}
fixed = list(gamma_mean = c(0.3, 0.25, 0.2))
m <- monty_dsl({
  alpha ~ Exponential(mean = 0.5)
  beta[, ] ~ Normal(0, 1)
  gamma[] ~ Exponential(mean = gamma_mean[i])
  
  dim(beta) <- c(2, 2)
  dim(gamma, gamma_mean) <- 3
}, fixed = fixed, gradient = FALSE)
```

we have two array parameters: `beta` (a 2x2 matrix) and `gamma` (a vector of length 3). Additionally, `gamma_mean` is a fixed vector of length 3.

As with arrays in `odin`, all arrays used in the `monty` DSL need a `dim()` equation.

Note the names of the parameters

```{r}
m$parameters
```

For indexing arrays in the DSL (again, the same as in `odin`) there is no use of `i`, `j` etc on LHS - indexing on the LHS is implicit. Support is for up to 8 dimensions, with index variables `i`, `j`, `k`, `l`, `i5`, `i6`, `i7`, `i8`. In our above example we had empty indices on the LHS, which means we will loop over the entire range of that dimension, so essentially

```r
beta[, ] ~ Normal(0, 1)
gamma[] ~ Exponential(mean = gamma_mean[i])
```

generates loops

```r
for (i in 1:2) {
  for (j in 1:2) {
    beta[i, j] ~ Normal(0, 1)
  }
}
for (i in 1:3) {
  gamma[i] ~ Exponential(mean = gamma_mean[i])
}

For flexibility dimensions can be softcoded and passed in via `fixed`

```{r}
fixed = list(gamma_mean = c(0.3, 0.25, 0.2),
             n_beta_1 = 2, n_beta_2 = 2, n_gamma = 3)
m <- monty_dsl({
  alpha ~ Exponential(mean = 0.5)
  beta[, ] ~ Normal(0, 1)
  gamma[] ~ Exponential(mean = gamma_mean[i])
  dim(beta) <- c(n_beta_1, n_beta_2)
  dim(gamma, gamma_mean) <- n_gamma
}, fixed = fixed, gradient = FALSE)
m
```

You can define arrays with multiline equations

```{r}
fixed = list(n_beta = 5)
m <- monty_dsl({
  beta[1] ~ Exponential(mean = 0.3)
  beta[2:n_beta] ~ Exponential(mean = beta[i - 1])
  dim(beta) <- n_beta
}, fixed = fixed, gradient = FALSE)
```

but collectively the equations need to define all elements of the array, and each element can only be defined once.

Elements of an array can depend upon other elements of the same array that are defined earlier. We can see this in our previous example

```r
beta[1] ~ Exponential(mean = 0.3)
beta[2:5] ~ Exponential(mean = beta[i - 1])
```
which becomes clearer when we think how the loop will be generated

```r
beta[1] ~ Exponential(mean = 0.3)
for (i in 2:5) {
  beta[i] ~ Exponential(mean = beta[i - 1])
}
```
